The impacts of park access on health outcomes: a spatial comprehensive approach

### 2018-12-13 manuscript.Rnw
Various data cleaning and visualization
Add a map illustrating the logsum value. Inspecting this figure encouraged me to change the distance calculation. We previously had tract centroid to park centroid; we now use population-weighted tract centroid to edge of park. For cases where the population-weighted centroid is *in* a park, we set the minimum distance to 0.1 miles to avoid infinite logsums.

\maketitle
<<setup, cache=FALSE, echo=FALSE, message=FALSE>>=
library(tidyverse)
library(geojsonio)
library(rgdal)
library(rgeos)
library(maptools)
library(spdep)
library(sf)
library(VGAM)
library(qwraps2)
library(knitr)
library(kableExtra)
opts_chunk$set(cache=TRUE)

@@

\section{Methodology}

<<tracts>>=
tracts <- geojson_read("data/nyc_tracts.geojson", what = "sp")
tracts_sf <- st_read("data/nyc_tracts.geojson", quiet = TRUE) %>%
  st_transform(4326)

# identify complete cases
complete_tracts_index <- tracts_sf %>%
  st_set_geometry(NULL) %>%
  tbl_df() %>%
  complete.cases()
tracts <- tracts[complete_tracts_index, ]

tracts@data <-  tbl_df(tracts@data) %>%
  dplyr::transmute(
    county_fips = substr(GEOID, 3, 5),
    borough = case_when(
      county_fips == "081" ~ "Queens",
      county_fips == "047" ~ "Brooklyn",
      county_fips == "061" ~ "Manhattan",
      county_fips == "005" ~ "Bronx",
      county_fips == "085" ~ "Staten Island",
      TRUE ~ as.character(NA)
    ),
    obesity = OBESITY,
    physact = 100 - Phys_Act,
    log_obesity = log(OBESITY),
    log_physact = log(100 - Phys_Act),
    density = Pop_Density,
    fulltime = FulltimeWork, college = CollegeDeg, 
    single = Single_Percent,
    black = PctBlack, asian = PctAsian, hispanic = PctHispanic,
    other = PctNative + PctPacific,
    Pct0to17, Pct18to29, Pct65plus, 
    Income1, Income2, Income3, Income4, Income5, Income6, 
    Income7, Income8, Income9, Income10
  )

tracts_sf <- st_as_sf(tracts)
@

We obtained data on aggregate health outcomes from the Centers for Disease
Control and Prevention (CDC) for census tracts in the New York City area. After
removing tracts with missing population information, we have Sexpr{nrow(tracts)}
complete cases.

\begin{table*}
  \caption{Summary Statistics of Tract Data}
  \label{tab:parkbenefits}
<<data_summary, results="asis", message=FALSE, warning=FALSE, cache=FALSE>>=
our_summary <- list(
  "obesity"     = tab_summary(tracts$obesity),
  "physact"     = tab_summary(tracts$physact),
  "pop_density" = tab_summary(tracts$density),
  "fulltime"    = tab_summary(tracts$fulltime),
  "college"     = tab_summary(tracts$college)
)

@@

<<parks>>=
parks <- sf::st_read("data/nyc_parks.geojson", quiet = TRUE) %>%
  mutate(
    log_size = log(Park_Acres),
    log_tweets = yeo.johnson(TWEET_COUNT, 0)
  )
@

@@

In the absence of a park choice survey, we estimated likely values of the $\beta$
coefficients by iteratively searching for the values which produced the highest
model likelihood in *which model did we use to estimate likelihood?*. The search
was constrained by two assertions. First, we required that the coefficient on
size be positive and that on distance negative; all else equal, people will
prefer to use larger parks that are nearer to their residence. Second, we visually
inspected the resulting accessibility scores to ensure that the scores produced a
reasonably varied pattern of access throughout the metropolitan region.

<<f_logsums>>=
#' Calculate destination choice logsums from a distance matrix
#' 
#' @param d An $n\times p$ matrix with the distance from all tracts to
#'   all parks
#' @param sizes A p-length vector of park sizes
#' @param tweets A p-length vector of tweets at parks
#' @param betas A vector containing the size, distance, and tweet coefficients. 
#'   If we submit two variables the tweets are ignored.
#'   
#' @return An n-length vector containing the weighted log-sum based
#'   accessibility between a tract and all parks.
#' @details If we have n tracts and p parks, distances needs to be a 
#' 
calculate_park_logsums <- function(d, sizes, tweets = NULL,
                                   betas = c(-15,.00001, .001)){

@@

\subsection{Spatial dependence and correlation}

Using tracts as the unit of observation in the model allows us to consider the
availability of parks at a metropolitan level, but it also introduces the problem
spatial correlation of unobserved errors and spatial dependence in the data
generating process \citep{Cliff1970, Anselin1980}. Spatial correlation occurs
when spatially-distributed unobservable or missing variables (school quality,
neighborhood prestige, etc.) influence the modeled outcomes. Spatial dependence
occurs when the outcome {\em depends} on the outcome in neighboring areas; for
example, seeing neighbors exercising may encourage exercise. The relaxed spatial
Durbin model (SDM) proposed by \citet{Burridge1981} controls for both processes:
\begin{equation}\label{eq:sdm}
  \vect{y}= \rho W\vect{y} + X\vect{\beta} + WX\vect{\gamma} + \vect{\epsilon}
\end{equation}
where $\rho$ is the estimated spatial correlation between the outcome variables
$y_i, y_j, \ldots  \in \vect{y}$, $W$ is a matrix specifying the spatial
relationship between observations $i$ and $j$ in the dataset,
$\vect{\beta}$ is a vector of estimable coefficients relating the attributes $X$
of an observation to its outcome, and $\vect{\gamma}$ is a vector of
estimated coefficients relating the attributes of an observation's {\em neighbors}
to its outcome.

It may in some cases be more efficient to estimate a restricted spatial autoregressive
model, that controls for only correlation or only dependence. \citet{tra_autoregression_paper}
illustrate a decision algorithm --- proposed by \citet{Florax2003} --- which uses
a likelihood ratio test of the SDM against a restricted spatial error model (SEM):
if the likelihood ratio test rejects that the models are equivalent, then the SDM
should be used.

@@

>>>>

\maketitle
<<setup, cache=FALSE, echo=FALSE, message=FALSE>>=
library(tidyverse)
library(geojsonio)
library(rgdal)
library(rgeos)
library(maptools)
library(spdep)
library(sf)
library(VGAM)
library(qwraps2)
library(knitr)
library(kableExtra)
opts_chunk$set(cache=TRUE)
library(leaflet)

@@

\section{Empirical Application}

\subsection{Data}

<<tracts>>=
tracts <- geojson_read("data/nyc_tracts.geojson", what = "sp")

# identify complete cases
complete_tracts_index <- tracts@data %>%
  tbl_df() %>%
  complete.cases()
tracts <- tracts[complete_tracts_index, ]

tracts@data <-  tbl_df(tracts@data) %>%
  dplyr::transmute(
    geoid = GEOID,
    county_fips = substr(GEOID, 3, 5),
    borough = case_when(
      county_fips == "081" ~ "Queens",
      county_fips == "047" ~ "Brooklyn",
      county_fips == "061" ~ "Manhattan",
      county_fips == "005" ~ "Bronx",
      county_fips == "085" ~ "Staten Island",
      TRUE ~ as.character(NA)
    ),
    obesity = OBESITY,
    physact = 100 - Phys_Act,
    log_obesity = log(OBESITY),
    log_physact = log(100 - Phys_Act),
    density = Pop_Density,
    fulltime = FulltimeWork, college = CollegeDeg,
    single = Single_Percent,
    black = PctBlack, asian = PctAsian, hispanic = PctHispanic,
    other = PctNative + PctPacific,
    Pct0to17, Pct18to29, Pct65plus,
    Income1, Income2, Income3, Income4, Income5, Income6,
    Income7, Income8, Income9, Income10
  ) 
@

This study used data available to the public from a variety of agencies and
data providers.
We obtained data on aggregate health outcomes from the Centers for Disease
Control and Prevention (CDC) for census tracts in five boroughs of New York City.
After removing tracts with missing population information, we have
Sexpr{nrow(tracts)} complete cases. Table \ref{tab:tractsdata} presents key
descriptive statistics for these data.

\begin{table*}
  \caption{Summary Statistics of Tract Data}
  \label{tab:tractsdata}
<<data_summary, results="asis", message=FALSE, warning=FALSE, cache=FALSE>>=
our_summary <- list(
  "obesity"     = tab_summary(tracts$obesity),
  "physact"     = tab_summary(tracts$physact),
  "pop_density" = tab_summary(tracts$density),
  "fulltime"    = tab_summary(tracts$fulltime),
  "college"     = tab_summary(tracts$college)
)

@@

<<parks>>=
parks <- sf::st_read("data/nyc_parks.geojson", quiet = TRUE) %>%
  transmute(
    size = Park_Acres,
    log_size = log(Park_Acres),
    tweets = TWEET_COUNT, 
    log_tweets = yeo.johnson(TWEET_COUNT, 0)
  )
@

@@

\subsection{Model}

We predict the obesity rate in a census tract as a function of the tract's
sociodemographic characteristics and choice-based accessibility to parks with a
semi-log specification:
\begin{equation}\label{eq:themodel}
\log(\mathit{Obesity})_i = f(\vect{X}_i, A_i, g(W, \mathit(Obesity), \vect{X}))
\end{equation}
<<base_formula>>=
base_formula <- formula( 
  ~ log(density) + 
    fulltime + college + single + 
    Pct0to17 + Pct18to29 + Pct65plus + # need to leave out a category for collinearity  
    black + asian + hispanic + other)
@
where $\vect{T}_i$ is a vector of tract-level attributes, $A_i$ is a choice-based
accessibility measure for the tract, and $g(W, \mathit(Obesity), \vect{T})$
represents the impact of spatial neighborhood effects on the outcome.
Using tracts as the unit of observation in the model allows us to consider the
availability of parks at a metropolitan level, but it also introduces the problem
of spatial correlation of unobserved errors and spatial dependence in the data
generating process \citep{Cliff1970, Anselin1980}. Spatial correlation occurs
when spatially-distributed unobservable or missing variables (school quality,
neighborhood prestige, etc.) influence the modeled outcomes. Spatial dependence
occurs when the outcome {\em depends} on the outcome in neighboring areas; for
example, seeing neighbors exercising may encourage exercise. The relaxed spatial
Durbin model (SDM) proposed by \citet{Burridge1981} controls for both processes:
\begin{equation}\label{eq:sdm}
  \vect{y}= \rho W\vect{y} + X\vect{\beta} + WX\vect{\gamma} + \vect{\epsilon}
\end{equation}
where $\rho$ is the estimated spatial correlation between the outcome variables
$y_i, y_j, \ldots  \in \vect{y}$, $W$ is a matrix specifying the spatial
relationship between observations $i$ and $j$ in the dataset,
$\vect{\beta}$ is a vector of estimable coefficients relating the attributes $X$
of an observation to its outcome, and $\vect{\gamma}$ is a vector of
estimated coefficients relating the attributes of an observation's {\em neighbors}
to its outcome.

It may  be more efficient to estimate a restricted spatial autoregressive
model -- that controls for only correlation or only dependence -- and therefore
save degrees of freedom in the estimation.  \citet{tra_autoregression_paper}
illustrate a decision algorithm originally proposed by \citet{Florax2003} which uses
a likelihood ratio test of the SDM against a restricted spatial error model (SEM):
if the likelihood ratio test rejects that the models are equivalent, then the SDM
should be used.

\subsection{Accessibility}

Extant destination choice models used in practice typically use travel time in
minutes as a travel impedance term, and employment by sector as a size term or
attraction component. Typically such models handle recreational trips together with
other non-work and non-school trips, and we therefore can find no previously
estimated coefficients using size terms relevant to park acreage or amenities.
Given this gap in the literature, we instead assert values for $\beta$ to generate
the choice-based accessibility statistic used in this study.

<<f_logsums>>=
#' Calculate destination choice logsums from a distance matrix
#'
#' @param d An $n\times p$ matrix with the distance from all tracts to
#'   all parks
#' @param sizes A p-length vector of park sizes
#' @param tweets A p-length vector of tweets at parks
#' @param betas A vector containing the size, distance, and tweet coefficients.
#'   If we submit two variables the tweets are ignored.
#'
#' @return An n-length vector containing the weighted log-sum based
#'   accessibility between a tract and all parks.
#' @details If we have n tracts and p parks, distances needs to be a
#'
calculate_park_logsums <- function(d, sizes, tweets = NULL,
                                   betas = c(-0.4,0.2,0.001)){

@@

We initially estimated potential values of the $\beta$ coefficients by
applying a limited-memory, box-constrained optimization algorithm maximizing the
SDM model log-likehood. The constraint required that the coefficient on
size be positive and that on distance negative; all else equal, we assert people
will prefer to use larger parks and parks that are nearer to their residence. This
resulted in $\beta_d = -0.42$  and $\beta_s = 0.000$. Given our expectation that
park size is a valuable attribute, we subsequently adjusted this value to

<<distances, warning=FALSE>>=
# calculate centroids of tracts and parks
pop_weighted_centroids <- read_csv("data/CenPop2010_Mean_TR36.csv") %>%
  mutate(geoid = str_c(STATEFP, COUNTYFP, TRACTCE))
pop_weighted_centroids <- left_join(tracts@data, pop_weighted_centroids, by = "geoid") %>%
  dplyr::select(geoid, LONGITUDE, LATITUDE) %>%
  st_as_sf(crs = 4326, coords = c("LONGITUDE", "LATITUDE")) %>%
  st_transform(2263) 
parks <- parks %>% st_transform(st_crs(pop_weighted_centroids))

# distances from all tracts to all zones
distances <- st_distance(pop_weighted_centroids, parks, byid = TRUE) %>%
  units::set_units(miles) %>%
  units::drop_units()

# assert that a tract must be at least 1/10 mile from a park
distances <- pmax(distances, 0.1)
@

<<logsum_betas>>=
tracts@data <- tracts@data %>%
  mutate(
    access_ls = calculate_park_logsums(log(distances), parks$log_size,
                                       betas = c(-0.42, 1.0)),
    tweets_ls = calculate_park_logsums(log(distances), parks$log_size, parks$log_tweets,
                                       betas = c(-0.42, 1.0)),
    distance_ls = (distance_ls - mean(distance_ls) )/ sd(distance_ls),
    access_ls = (access_ls - mean(access_ls) )/ sd(access_ls),
    tweets_ls = (tweets_ls - mean(tweets_ls) )/ sd(tweets_ls)
  )

tracts_sf <- tracts %>% st_as_sf() %>% st_transform(4326)
@

<<leaflet, eval=FALSE>>=
# this interactive map is much easier to create and explore
pal <- colorQuantile("Spectral", tracts_sf$access_ls, n = 5)
leaflet(tracts_sf) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(group = "access", color = ~ pal(access_ls), label = ~as.character(round(access_ls,2))) 
@

\begin{figure*}

<<map>>=
map <- ggmap(get_map(unname(st_bbox(tracts_sf)), source = "stamen", maptype = "toner"),
             extent = "device")
map +
  geom_sf(data = tracts_sf, inherit.aes=FALSE, 
          aes(fill = cut(access_ls, quantile(access_ls))), alpha = 0.8) +
  scale_fill_brewer("Relative Park Logsum", palette = "GnBu")
@
  \caption{Relative surplus-based accessibility in New York City}
  \label{fig:map}
\end{figure*}

@@

We estimate

```{r spatial selection}
obese_base_lm <- lm(update(base_formula, log_obesity ~ .), data = tracts@data)
obese_base_sar <- lagsarlm(update(base_formula, log_obesity ~ .), 
                           data = tracts@data, listw = W, zero.policy = TRUE)
obese_base_sem <- errorsarlm(update(base_formula, log_obesity ~ .), 
                             data = tracts@data, listw = W, zero.policy = TRUE)
obese_base_sdm <- lagsarlm(update(base_formula, log_obesity ~ .), 
                           data = tracts@data, listw = W, zero.policy = TRUE, 
                           type = "mixed")
```

A likelihood ratio test reveals that the SEM is not preferred, so we use the SDM
only going forward.

```{r lrtest}
lmtest::lrtest(obese_base_sem, obese_base_sdm)
```

```{r obesity_access_models}
access_sdm <- update(obese_base_sdm, .~ . + access_ls)
tweets_sdm <- update(obese_base_sdm, .~ . + tweets_ls)
```
